Layered Formal Verification of a TCP Stack

### 2021-06-03 SecDev21/paper.tex
Rewording, add abstract + contributions

\title{Layered Formal Verification of a TCP/IP Stack}

@@

\begin{abstract}
\todo{Write the abstract.}
\end{abstract}

@@

In total, for this work, 50 subprograms have been translated from C to SPARK. This accounts for 2266 lines of code, of which 1165 are SPARK logic code (e.g.\ contracts) used to represent and prove the protocol's specification. The overall proving time with GNATprove is 9m and 54s on a 12-processor Intel(R) Core(TM) i7-9750H CPU with 2.60GHz speed and 16GB of RAM.

%\todo[inline]{KG:pending contributions list and fixing the next paragraph. Will be fixed when the rest of the paper is finalized.}

@@

TCP is based on a multitasking model. Different tasks can interact and update the socket data structure, used to retain the status of a TCP connection, to handle the various events that can occur within a TCP session. RFC 793, under Section ``3.9. Event Processing'' page 52, describes a possible implementation of how to handle these events based on three tasks: one for the \emph{user calls}, one for the \emph{arriving segments} and one for the \emph{timers}. The role of each of the three tasks can be summarized as follows. User calls refer to functions, namely \texttt{open}, \texttt{close}, \texttt{abort}, \texttt{send} and \texttt{receive}, that can be called by the user to control the connection, send, or receive data. These functions can trigger transitions between the connection's states since they are intended to control the connection. In the arriving segments task, the received segments are being processed,
and the corresponding messages are sent back. Transitions between states can be triggered on the reception of a segment. Finally, timers control the timeouts, for example, the retransmission timeout to retransmit a message or the time-wait timeout to close the connection after a specified amount of time elapsed. Thus, transitions corresponding to the timeout events can also be triggered by this task. A complete description of the protocol's specification can be found in the RFC 793 document~\cite{rfc793}.

@@

Symbolic execution is used to verify that all the possible paths when executing the \spark{tcpProcessSegment} function will result in a state that respects the RFC 793 extracted specifications. This process requires three steps shown in \Cref{code:kleedriver}. Firstly, a symbolic incoming segment is created, allowing to account for all possible incoming segments (see lines 1-4). Then at \cref{code:kleedriver:tcpProcessSegment}, function \texttt{tcpProcessSegment} is executed symbolically by KLEE. Finally, the assertion, \texttt{klee\_assert}, at \cref{code:kleedriver:assert} checks if the code leads to a valid state, one that is within the protocol's specification. If there is no raised assertion, then the contract represented by the assertion is deemed proved, and the relevant C code respects the protocol's specification regarding that contract. The assertion is then converted to a SPARK contract and added to the function \spark{TCP_Process_One_Segment}. This enables the SPARK prover to reason about the possible side-effects caused by the C-code on the state of a TCP connection and thus, enables the functional correctness verification of the TCP user functions.

@@

@@

The pointer ownership policy of SPARK, detailed in \Cref{sec:pointers}, enabled the detection of a memory leak when a socket closes. This was present due to partially freeing the memory allocated for the buffer that stores the incoming segments. This bug was due to using the wrong library function call to free the memory. SPARK ownership policy detected this because, after the call to this function, the pointer to the buffer was declared not own anymore, but the memory pointed by that pointer was partially deallocated.

@@

A subtle bug that violates the protocol's extracted specifications captured by SPARK is reproduced below:

@@

In the original implementation, after the call to \spark{Tcp_Send}, if no error occurs, the state of the socket can either be \sestab or \sclosew. After the call at line~\ref{bugProg:tcpWaitForEvents}, the state of the socket can still be \sestab or \sclosew, or \sclosed if a \rst flag has been received when the mutex was released. At line~\ref{bugProg:tcpWaitForEventsPost}, a \fin segment is sent, which does not modify the state of the socket. Finally, at line~\ref{bugProg:tcpChangeState} the code tries to move the connection to the \sfwone state, and thus, the transitions $\text{\sclosew} \rightarrow \text{\sfwone}$ and $\text{\sclosed} \rightarrow \text{\sfwone}$ are possible.

These transitions are not allowed by RFC 793, and thus, not allowed by the precondition of \spark{TCP_Change_State}, which models all the allowed transitions by the protocol's specification. The problem was both corrected in the C original implementation and the SPARK translated code.

% The above is uncler, the transitions from CLOSED and CLOSED-WAIT are allowed, it's only the transition from ESTABLISHED which is not allowed.

>>>>

\title{Layered Formal Verification of a TCP Stack}

@@

\begin{abstract}
The Transmission Control Protocol (TCP) at the heart of TCP/IP protocol stacks
is a critical part of our current digital infrastructure. In this article, we
show how an existing professional-grade open source embedded TCP/IP library can
benefit from a formally verified TCP reimplementation. Our approach is to apply
formal verification to the TCP layer only, relying on validated models of the
lower layers on which it depends.
\end{abstract}

@@

In total, for this work, 50 subprograms have been translated from C to SPARK. This accounts for 2266 lines of code, of which 1165 are SPARK logic code (e.g.\ contracts) used to represent and prove TCP specification. The overall proving time with GNATprove is 10 minutes on a 12-processor Intel(R) Core(TM) i7-9750H CPU with 2.60GHz speed and 16GB of RAM.

The contributions of this article are the following. First, we show that formal verification can be applied selectively to a layer of an existing TCP/IP library, by translating the TCP layer to SPARK and modelling the lower layers on which it depends. The benefits are three-fold: we found two bugs in the TCP layer, we proved the correction of that layer, and we defined formal contracts for the upper layers to call the TCP layer. Second, we showed how the models of the lower layers can be validated by symbolic execution with KLEE on the C code of those layers.

@@

TCP is based on a multitasking model. Different tasks can interact and update the socket data structure, used to retain the status of a TCP connection, to handle the various events that can occur within a TCP session. RFC 793, under Section ``3.9. Event Processing'' page 52, describes a possible implementation of how to handle these events based on three tasks: one for the \emph{user calls}, one for the \emph{arriving segments} and one for the \emph{timers}. The role of each of the three tasks can be summarized as follows. User calls refer to functions, namely \texttt{open}, \texttt{close}, \texttt{abort}, \texttt{send} and \texttt{receive}, that can be called by the user to control the connection, send, or receive data. These functions can trigger transitions between the connection's states since they are intended to control the connection. In the arriving segments task, the received segments are being processed,
and the corresponding messages are sent back. Transitions between states can be triggered on the reception of a segment. Finally, timers control the timeouts, for example, the retransmission timeout to retransmit a message or the time-wait timeout to close the connection after a specified amount of time elapsed. Thus, transitions corresponding to the timeout events can also be triggered by this task. A complete description of TCP specification can be found in the RFC 793 document~\cite{rfc793}.

@@

Symbolic execution is used to verify that all the possible paths when executing the \spark{tcpProcessSegment} function will result in a state that respects the RFC 793 extracted specifications. This process requires three steps shown in \Cref{code:kleedriver}. Firstly, a symbolic incoming segment is created, allowing to account for all possible incoming segments (see lines 1-4). Then at \cref{code:kleedriver:tcpProcessSegment}, function \texttt{tcpProcessSegment} is executed symbolically by KLEE. Finally, the assertion, \texttt{klee\_assert}, at \cref{code:kleedriver:assert} checks if the code leads to a valid state, one that is within TCP specification. If there is no raised assertion, then the contract represented by the assertion is deemed proved, and the relevant C code respects TCP specification regarding that contract. The assertion is then converted to a SPARK contract and added to the function \spark{TCP_Process_One_Segment}. This enables the SPARK prover to reason about the possible side-effects caused by the C-code on the state of a TCP connection and thus, enables the functional correctness verification of the TCP user functions.

@@

As a result of this work, two bugs were detected and reported to the authors of
CycloneTCP, and they fixed them in the C implementation.

@@

A memory leak was detected thanks to the pointer ownership policy of SPARK
detailed in \Cref{sec:pointers}. When closing a socket, the associated memory
allocated for the buffer that stores incoming segments was only partially
freed, due to the use of the wrong library function to free the memory.

@@

A subtle bug violating TCP specification is reproduced below:

@@

In the original implementation, after the call to \spark{Tcp_Send}, if no error occurs, the state of the socket can either be \sestab or \sclosew. After the call at line~\ref{bugProg:tcpWaitForEvents}, the state of the socket can still be \sestab or \sclosew, or \sclosed if a \rst flag has been received when the mutex was released. At line~\ref{bugProg:tcpWaitForEventsPost}, a \fin segment is sent, which does not modify the state of the socket. Finally, at line~\ref{bugProg:tcpChangeState} the code tries to move the connection to the \sfwone state, and thus, the transitions $\text{\sclosew} \rightarrow \text{\sfwone}$ and $\text{\sclosed} \rightarrow \text{\sfwone}$ are possible.
These transitions are not allowed by RFC 793, and thus, not allowed by the precondition of \spark{TCP_Change_State}, which models all the transitions allowed by TCP specification. Thus GNATprove reported an unproved precondition check here.
